---
output: github_document
---

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/CFIA-PDAM/protoclim/graph/badge.svg)](https://app.codecov.io/gh/CFIA-PDAM/protoclim)
[![R-CMD-check](https://github.com/CFIA-PDAM/protoclim/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CFIA-PDAM/protoclim/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

devtools::install()
library(dplyr)
library(protoclim)
```

# protoclim

## Overview

`{protoclim}` is an R package for determining the climatic suitability of plant species and is intended to support risk assessors in performing a formal [pest risk analysis](https://www.ippc.int/en/publications/592/). `{protoclim}` is heavily inspired and influenced by the [Proto3](https://link.springer.com/article/10.1007/s10530-017-1567-1) technique developed and used by the United States Department of Agriculture in their [Guidelines for the USDA-APHIS PPQ Weed Risk Assessment Process](https://www.aphis.usda.gov/sites/default/files/wra-guidelines.pdf). 

`{protoclim}` is designed to be interoperable with any categorical climate variable but being inspired by the Proto3 model is intended to be used with: 

- Köppen-Geiger classes
- Precipitation bands
- Plant hardiness zones

Users *are* able to bring their own data but high-resolution global data has already been currated for each of the above climate variables by the associated repository [`protoclimData`](https://github.com/CFIA-PDAM/protoclim) and is available for download from [Zenodo](https://zenodo.org/records/21400412). The climate variables are available across present-day and future time periods as well as various emission scenarios. 

In addition to climate data, using `{protoclim}` requires species occurrence data from GBIF, the Global Biodiversity Information Facility. GBIF occurrence data can be downloaded from the [GBIF website](https://www.gbif.org/occurrence/search?occurrenceStatus=present), the `{rgbif}` R package, or various other means. 

`{protoclim}` comes with two datasets: 

  1. `bryonia_alba`, GBIF occurrence data for *Bryonia alba*, the white bryony
   
  2. `climate_layers_ssp370_2011_2040`, a lower-resolution version of the associated `.tif` file hosted on [Zenodo](https://zenodo.org/records/21400412). 

```{r}
head(bryonia_alba)
```

```{r}
# Note that a function is used to read the climate data
climate_layers_ssp370_2011_2040()
```

## Installation

The easiest way to get `{protoclim}` is to install it from CRAN:

```{r, eval = FALSE}
install.packages("protoclim")
```

### Development version

To get a bug fix or to use a feature from the development version, you can install the development version of protoclim from GitHub:

```{r, eval = FALSE}
# install.packages("pak")
pak::pkg_install("CFIA-PDAM/protoclim")
```

## Usage




## Getting help

If you encounter a clear bug, please file an issue with a minimal reproducible example on [Github](https://github.com/CFIA-PDAM/protoclim/issues).