---
title: "Example workflow"
execute:
  freeze: auto
format:
  html:
    lightbox: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Researchers, botanists, and those working at National Plant Protection Organization's (NPPOs) often need to assess the climatic suitability of a potential invasive plant species in a given region. The workflow illustrated here shows how to use the `{protoclim}` package to assess the climatic suitability of _Bryonia alba_, the [eastern white-bryony](https://www.gbif.org/taxon/NJ97), in North America. 

Two types of data are required to assess the climatic suitability of a plant species with `{protoclim}`:

1. A data frame of biological occurrence data from GBIF. 

2. A global raster of Köppen-Geiger climate classes, plant hardiness zones, and precipitation bands. 

Available within `{protoclim}` is the `bryonia_alba` dataset, which contains GBIF occurrence data for `Bryonia alba`, as well as the helper function `climate_layers_ssp370_2011_2040()` which returns a downscaled global raster of Köppen-Geiger climate classes, plant hardiness zones, and precipitation bands for SSP 3.7.0 for 2011-2040. 

::: {.callout-tip}

The climate data provided by `climate_layers_ssp370_2011_2040()` is relatively low resolution and only covers one emission scenario and time period. Higher resolution climate data for various emission scenarios and time periods are pre-compiled and available for download on Zenodo at [10.5281/zenodo.21400410](https://zenodo.org/records/21400410).

:::

To start, load `{protoclim}` into your R session. 
```{r setup}
#| output: false
#| warning: false
library(protoclim)
```

Then look at `bryonia_alba` which contains GBIF occurrence data for _Bryonia alba_. 
```{r}
dplyr::glimpse(bryonia_alba)
```

Then look at the raster of climate data packaed with `{protoclim}` that comes with layers for Köppen-Geiger climate classes, precipitation bands, and plant hardiness zones. Each layer is the mode taken across multiple global climate models. See the [`protoclimData`](https://github.com/CFIA-PDAM/protoclimData) repository on GitHub for additional details.
```{r}
climate_layers_ssp370_2011_2040()
```

These two pieces of data are all that is needed to perform climate suitability analysis with `{protoclim}`. In your own work, the `{rgbif}` package makes extracting biological occurrence data from GBIF straightforward and the pre-compiled, high-resolution climate data available on [10.5281/zenodo.21400410](https://zenodo.org/records/21400410) means you will not need to spend time and effort compiling climate data yourself.

The challenge that remains is in cleaning the biological occurrence data. Packages like `{CoordinateCleaner}` can help to scale and automate this process but, depending on your use case, additional cleaning may be required. Here, we use `{CoordinateCleaner}` to clean the `bryonia_alba` occurrence data. See [Cleaning GBIF data for the use in biogeography](https://docs.ropensci.org/CoordinateCleaner/articles/Cleaning_GBIF_data_with_CoordinateCleaner.html) for additional information.
 
```{r}
#| output: false
#| warning: false

# Remove all occurrence records without coordinates
bryonia_alba <- dplyr::filter(
  bryonia_alba,
  !is.na(decimalLatitude),
  !is.na(decimalLongitude)
)

# Convert the ISO 2 country codes to ISO 3 country codes for use with CoordinateCleaner
bryonia_alba <- dplyr::mutate(
  bryonia_alba,
  countryCode = countrycode::countrycode(
    countryCode,
    origin = 'iso2c',
    destination = 'iso3c'
  )
)

# Clean the occurrence data using CoordinateCleaner
bryonia_alba <- CoordinateCleaner::clean_coordinates(
  bryonia_alba,
  value = "clean"
)
```

With the cleaned occurrence data `bryonia_alba` in hand, we can predict whether _Bryonia alba_ is climatically suitability in each level of the Köppen-Geiger climate classes, precipitation bands, and plant hardiness zones. The output of `classify_layers` is a list of tibbles, one per climate layer. Looking at the output below, we see that _Bryonia alba_ is predicted to be `Not suitable` in Köppen-Geiger climate classes 1 through 4, `Not likley to be suitable` in 5, and `Likely to be suitable` in 8. 

```{r}
(layer_classes <- classify_layers(
  climate_layers_ssp370_2011_2040(),
  bryonia_alba
))
```

With the `layer_classes` classifying the levels of each climate layer as `Suitable` through `Not suitable`, we return to the raster of climate data and determine whether each cell's level of suitability for _Bryonia alba_. We also take the minimum level of suitability across each climate layer in the `overall` layer. 

```{r}
(cell_classes <- classify_cells(
  climate_layers_ssp370_2011_2040(),
  layer_classes
))
```

```{r}
cell_classes <- terra::crop(
  cell_classes,
  geodata::gadm(country = "CAN", level = 0, path = tempdir()),
  mask = TRUE
)

cell_classes <- terra::project(cell_classes, "EPSG:102001", method = "near")

df <- as.data.frame(cell_classes$overall, xy = TRUE, na.rm = TRUE)

ggplot2::ggplot() +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(x = x, y = y, fill = overall),
    data = df,
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "Suitable" = "#FF0000",
      "Likely to be suitable" = "#FF4040",
      "Not likely to be suitable" = "#FF8080",
      "Not suitable" = "#FFBFBF"
    )
  ) +
  ggplot2::theme_void()
``` 